-------------------------------------------------------------------------------
# 1 Plot simple vectors

In this chapter we explore some of the graphical functionality of **R**. We first define 3 simple vectors and try to plot them.

```{r R1}
rm(list=ls()) # Remove almost everything in the memory
setwd("C:/Dropbox/Towson/Teaching/3_ComputationalEconomics/Lectures/Lecture2/R")
cars   = c(1, 3, 6, 4, 9)
trucks = c(2, 5, 4, 5, 12)
suvs   = c(4, 4, 6, 6, 16)
```

-------------------------------------------------------------------------------
## 1.1 First plot

We then simply plot the first vector using the ```plot()``` command.

```{r R2}
plot(cars)
```
We can change the way the graph looks adding plot symbols and color. In addition we add a title.

```{r R3}
# Graph cars using blue points overlayed by a line 
plot(cars, type="o", col="blue")
   # Create a title with a red, bold/italic font
   title(main="Autos", col.main="red", font.main=4)
```

-------------------------------------------------------------------------------
## 1.2 Adjusting the plot
We can also adjust the plot range in both dimensions as follows.

```{r R4}
plot(cars, type="o", col="blue", ylim=c(0,12)) # Create a title with a red, bold/italic font
   title(main="Autos", col.main="red", font.main=4)
```

-------------------------------------------------------------------------------
## 1.3 Graph autos: 2 types
Finally, we can the second vector to the plot using the command ```line```.
We graph autos using y axis that ranges from 0 to max value in cars or trucks vector.  
Turn off axes and annotations (axis labels) so we can specify them ourself.

```{r R5}
# Calculate range from 0 to max value of cars and trucks
g_range <- range(0, cars, trucks)

plot(cars, type="o", col="blue", ylim=g_range, 
     axes=FALSE, ann=FALSE)
   # Make x axis using Mon-Fri labels
   axis(1, at=1:5, lab=c("Mon","Tue","Wed","Thu","Fri"))
   
   # Make y axis with horizontal labels that display ticks at 
   # every 4 marks. 4*0:g_range[2] is equivalent to c(0,4,8,12).
   axis(2, las=1, at=4*0:g_range[2])
   
   # Create box around plot
   box()
   
   # Graph trucks with red dashed line and square points
   lines(trucks, type="o", pch=22, lty=2, col="red")
   
   # Create a title with a red, bold/italic font
   title(main="Autos", col.main="red", font.main=4)
   
   # Label the x and y axes with dark green text
   title(xlab="Days", col.lab=rgb(0,0.5,0))
   title(ylab="Total", col.lab=rgb(0,0.5,0))
   
   # Create a legend at (1, g_range[2]) that is slightly smaller 
   # (cex) and uses the same line colors and points used by 
   # the actual plots 
   legend(1, g_range[2], c("cars","trucks"), cex=0.8, 
          col=c("blue","red"), pch=21:22, lty=1:2);
```

-------------------------------------------------------------------------------
## 1.4 Graph autos: 3 types

Finally, we graph all three into one figure.
This time we save the graph as ```fig1.pdf``` into subfolder Graphs.

```{r R6}
max_y = max(cars, trucks, suvs)

# Define colors to be used for cars, trucks, suvs
plot_colors = c("blue","red","forestgreen")

# Start PDF device driver to save output to fig1.pdf
# pdf(file="./Graphs/fig1.pdf")

   # Graph autos using y axis that ranges from 0 to max_y.
   # Turn off axes and annotations (axis labels) so we can 
   # specify them ourself
   plot(cars, type="o", col=plot_colors[1], 
        ylim=c(0,max_y), axes=FALSE, ann=FALSE)
   
      # Make x axis using Mon-Fri labels
      axis(1, at=1:5, lab=c("Mon", "Tue", "Wed", "Thu", "Fri"))
      
      # Make y axis with horizontal labels that display ticks at 
      # every 4 marks. 4*0:max_y is equivalent to c(0,4,8,12).
      axis(2, las=1, at=4*0:max_y)
      
      # Create box around plot
      box()
      
      # Graph trucks with red dashed line and square points
      lines(trucks, type="o", pch=22, lty=2,col=plot_colors[2])
      
      # Graph suvs with green dotted line and diamond points
      lines(suvs, type="o", pch=23, lty=3,col=plot_colors[3])
      
      # Create a title with a red, bold/italic font
      title(main="Autos", col.main="red", font.main=4)
      
      # Label the x and y axes with dark green text
      title(xlab= "Days", col.lab=rgb(0,0.5,0))
      title(ylab= "Total", col.lab=rgb(0,0.5,0))
      
      # Create a legend at (1, max_y) that is slightly smaller 
      # (cex) and uses the same line colors and points used by 
      # the actual plots
      legend(1, max_y, c("cars","trucks","suvs"), cex=0.8, col=plot_colors, 
             pch=21:23, lty=1:3);    
# Turn off device driver (to flush output to png)
dev.off()
```


-------------------------------------------------------------------------------
# 2 Plotting functions


We next want to plot mathematical functions. We again first define the input vector ```x``` and then specify the "output" vector ```y``` as: $$y = f(x)$$ where for our first example $latex f(x) = x^2$ , $latex f(x)=3*y/2 - 5$, and finally $latex f(x)=5*y/2 - \sqrt{y}$.

```{r R7}
x = seq(1, 10, by=0.1)
y = x^2
y1 = 3*y/2 - 5
y2 = 5*y/2 - sqrt(y)
```

We have now two columns of values: x and y that we can plot against each other into a coordinate system.

```{r R8}
plot(x,y)
# plot(x,y,type="l")
   # if you want to add plots to the one we just made, use the 'line' command.
   lines(x,y1)
   lines(x,y2)
```


--------------------------------------------------------------------------------
# 3 3-D Graphs 

--------------------------------------------------------------------------------

## 3.1 Simply use x,y,z as coordinates for points in 3D space
Finally, we can also plot 3-D graphs in **R**. In order to do this we need the library ```rgl```.
This introduces a new command called ```plot3d``` with a slightly more involved syntax.

```{r R9}
library("rgl")

x = seq(1, 10, by=1)
y = seq(1, 10, by=1)
n = length(x)
z = 3*y^2/2 + x - sqrt(x*y)/ 5

# Simply use x,y,z as coordinates for points in 3D space
plot3d(x,y,z,type="p", col="red", xlab="X", ylab="Y", zlab="Z", 
       size=5, lwd=15, box=F)
```
--------------------------------------------------------------------------------
## 3.2 Use 'wireframe' command if you want to plot the graph over its entire domain

```{r R10}
library("lattice")
## Make data for wireframe command
i = 1
for (x0 in seq(1, 10, 1)) {
   for (y0 in seq(1, 10, 1)) {
     x[i] = x0
     y[i] = y0
     z[i] =  3*y[i]^2/2 + x[i] - sqrt(x[i]*y[i])/ 5
     i = i + 1  # data counter
   }
}
wireData = data.frame(x=x,y=y,z=z)

wireframe(z ~ x*y, main="3 D Wireframe Plot", xlab = "x", ylab = "y",
data=wireData)
```

--------------------------------------------------------------------------------
# 4 Subplots


If we have more than one figure it might be good to put them all into one graph. In the following example we plot 6 figures into one picture. We plot into 3 rows and 2 columns. You can obviously rearrange all this.

```{r R11}
 #  Open a new default device.
get(getOption("device"))()

 ##  Set up plotting in 3rows and 2 columns, plotting along rows first.
par(mfrow = c(3, 2))

plot(c(1,2,3), type="o")  
title('figure 1')


plot(runif(12), runif(12), type="l",col="red")
title('figure 2')


plot(c(1,1,1))  # implicitly creates subplot(111)
title('figure 3', pch=22, lty=2, col="red")


plot(runif(12), runif(12),  type="o", pch=23, lty=3, col="blue")
title('figure 4')


plot(c(1,2,1), type="l")  # implicitly creates subplot(111)
title('figure 5')

plot(runif(12), runif(12), type="l", lty=2, lwd=2)
title('figure 6')
```

------------------------------------------------------------------------------- 
# Exercises

1. Define vector $latex x=[0,...,10]$ and function y as $latex y(x) = \frac{\sqrt{x}}{x}$.
2. Plot values of $latex x$ against $latex y(x)$.
3. Define a new function $latex z(y) = y +5.5$ and plot it as subplot,
   underneath the first plot. Note that $latex z$ is ultimately a function of $latex x$, as in $latex z(y(x))$.
4. Add a third subplot that has both, $latex y$ and $latex z$ in the same
   graph.
